A family of accessibility measures

### 2024-11-07 paper/manuscript.qmd
Document after discussion with Anastasia

---
title: "A family of accessibility measures"
format: pdf
bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
nocite: |
  @careyPrinciplesSocialScience1858, @ravensteinLawsMigration1889, @ravensteinLawsMigration1885
---

@@

<!-- SUMMARY: Accessibility is related to mobility, however, institutionalized mobility-based transportation planning practice has treated accessibility as it's proxy. This has had problematic consequences. Conceptually, accessibility is a more complete and comprehensive evolution of mobility. As such, there has been recognition that the adoption of accessibility planning approaches are needed; however there are 1) too many definitions and 2) uncertainty in meaning of the outputs--> 
Accessibility is conceptually related to human mobility, it is often defined as the "potential of opportunities for interaction" [@hansen1959], opportunities such as places, events and social contacts that are important to people [@ferreiraReenactingMobilityAccessibility2020]. Accessibility adds a more comprehensive and complete understanding onto mobility, which simplify emphasizes the ability to move. However, the motivation of transportation planning has historically centered mobility and treated accessibility as its proxy, leading to problematic consequences [@handy2020]. In the rise of automobility-dominance, recent decades of mobility-based practice (e.g., highway expansion) have been marked by increased travel cost (time, effort, energy, financial cost) often not having an impact on accessibility levels [@handyACCESSIBILITYVSMOBILITYENHANCING2002]. In response, transportation planning practice has begin to adopt accessibility planning approaches, be it through the use of accessibility planning tools [@silvaAccessibilityInstrumentsPlanning2017] or through the traditional four-step transportation demand model [@ortuzar_2011_modelling]. However, institutional and conceptual barriers remain, many of which relate to the diversity of accessibility definitions [@handy2020] and tangible meaning of the outputs [@ferreiraReenactingMobilityAccessibility2020].

@@

<!-- PAPER'S AIM: improve interpretation of accessibility measure outputs. We do this by, 1) assuming accessibility is more complete understanding of mobiltiy, and connected them both mobility and accessibility are both based in the Netownian gravitational anology, lets introduce balancing factors like those  Gravity Model.   -->
In adopting the perspective that accessibility is an evolution of mobility conceptually, this paper aims to re-unify accessibility measures with its gravity-based Newtonian roots. It proposes to do so by mapping accessibility onto the family of gravity models. Not only to clarify accessibility's interpretation as a measure, but also to encourage future developments that draw from its physics roots, gravity modelling siblings, and spatial interaction history. We began by tracing the development of spatial interaction from @ravensteinLawsMigration1889 to it's split into accessibility research from @hansen1959 to today. In the following section, we will derive different types of accessibility measures, following the family of gravity models namely: non-constrained, constrained, unconstrained/Total constrained, singly constrained, composite singly constrained, doubly constrained. Then, we will provide empirical examples of the types of measures. Finally, we'll discuss the use of each measure and broadly the implications of ignoring the issue of proportionality in accessibility analysis.

@@

## The inquiry into spatial interaction traces to Newton's Universal Gravitation
The patterns of people's movement in space have been a subject of scientific inquiry for centuries. From as far back as 1858 in Henry C. Carey's _Principles of Social Science_ where "man [is] the molecule of society [and their interaction is subject to] the direct ratio of the mass and the inverse one of distance" [@mckeanManual1883, pp. 37-38], investigations of how people interact in space has often been explicitly coloured by the features of Newton's Law of Universal Gravitation [posited in _Principia Mathematica_ of 1687 and expressed in @eq-grav-prop].

$$
F_{ij} \propto \frac{M_i M_j} {D_{ij}^{2}}
$${#eq-grav-prop}

@eq-grav-prop is one of the most famous equations in all of science, and it states that the force of attraction $F$ between a pair of bodies $i$ and $j$ is directly _proportional_ to the product of their masses $M_i$ and $M_j$, and inversely _proportional_ to the square of the distance between them $D_{ij}$. Direct proportionality only means that as the product of the masses increases, so does the force. Likewise, inversely proportional only means that as the distance increases, the force decreases. To quantify the magnitude of the force, an empirical factor is required to convert the proportionality into an equation, which ensures that the force $F$ matches empirical observations of the force of attraction between masses. Ultimately, the equation for the force is as seen in @eq-grav, where $G$ is a proportionality constant obtained empirically:

$$
F_{ij} = G \frac{M_i M_j} {D_{ij}^{2}}
$${#eq-grav}

## Early human spatial interaction research: from Ravenstein 1889 to Stewart 1948
<!-- SUMMARY: from the 1880s to the 1940s, researchers have theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance. This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy. -->
Following Carey, research into human spatial interaction continued in different contexts. In 1885 and 1889, Ravenstein proposed his "Laws of Migration", based on his empirical analysis of migration flows in various countries [@ravensteinLawsMigration1885;@ravensteinLawsMigration1889]. In these works, Ravenstein posited a directly proportional relationship with the size of the destination (i.e., centres of commerce and industry), and an inversely proportional relationship between the size of flows and separation between origin and destination. These propositions are similar to Newton's gravitational laws. Over time, other researchers discovered similar relationships. <!--Rilley does not cite Newton, nor Carey, nor Ravenstein, but I removed this curiosity as it's not central to our point here--> For example, @reilly1929methods formulated a law of retail gravitation, expressed in terms of equal attraction to competing retail destinations. Later, Zipf proposed a $\frac{P_1P_2}{D}$ hypothesis for the case of goods movement by railways @zipfHypothesisCaseRailway1946, intercity personal movement @zipfHypothesisIntercityMovement1946,and information @zipfDeterminantsCirculationInformation1946. The $\frac{P_1P_2}{D}$ hypothesis stated that the magnitude of flows was proportional to the product of the populations of settlements, and inversely proportional to the distance between them.

A common theme among these early researchers of human spatial interaction is that a proportionality constant similar to $G$ in @eq-grav is missing. Of the researchers cited, only Reilly and Zipf expressed their hypotheses in mathematical terms. Reilly's hypothesis was expressed as follows:
$$
B_a = \frac{(P_a\cdot P_T)^N}{D_{aT}^n}
$${#eq-reilly}

@@

After Carey, it is in Stewart's [-@stewartDemographicGravitationEvidence1948] work on the principles of demographic gravitation that we find the strongest connection yet to Newton's law. We suspect that may relate to academic background; Stewart was a physicist, whereas Ravenstein, Reilly, and Zipf were social scientists. Besides awareness of preceding research, as he cites Reilly and Zipf as predecessors in the analysis of human spatial interaction, Stewart appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, as follows:

$$
F = G\frac{(N_1\mu_1)(N_2\mu_2)}{d_{12}^2}
$${#eq-stewart-force}

@@

Stewart's formulation of demographic force, developed in the context of what he called "social physics" [@stewartPrinciples1947], was problematic. It had issues with inconsistent mathematical notation. For example, e.g., in @stewartDemographicGravitationEvidence1948, $G$ is used as a proportionality constant for both _demographic force_ and  _demographic energy_ on p. 34 <!--Antonio, is this what you meant?-->. And perhaps more seriously, it was permeated by a view of humans as particles, following physical laws, laws of which were afflicted by unmitigated racism. For instance, molecular weight $\mu$ was taken as one for the average American, but "presumably...much less than one....for an Australian aborigine" [p. 35]. Stewart's ideas about "social physics" fell out of favour among social scientists, but not before influencing the field of accessibility today.

In summary: from the 1880s to the 1940s, researchers have theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $M$ of a pair of locations ($i$ and $j$) and inversely proportional by some sort of separation distance $D_{ij}$. Those that captured the relationship empirically, approached the concept with different expressions, some including a proportionality constant and some not, but all tying back to the Newtonian gravity analogy.

@@

From @stewardDemographicGravitationEvidence1948, we arrive to Walter G. Hansen, the author of @hansen1959 which has been exceptionally influential in the accessibility literature. In this seminal paper @hansen1959, "accessibility is defined as the potential of opportunities for interaction" [p.73], it is "a generalization of the population-over-distance relationship or "population potential" concept developed by Stewart [@stewardDemographicGravitationEvidence1948]" [p.73], explicitly defined "as the potential of opportunities for interaction" [p.73]. As well as being a student of city and regional planning at Massachusetts Institute of Technology, Hansen was also an engineer with the the Bureau of Roads likely preoccupied with applied research that can be used to qualify transportation's power to shape land uses. @hansen1959 focused on @stewartDemographicGravitationEvidence1948's _population potential_ (expressed in (@eq-stewart-population-potential-sum)), leaving out Stewart's other formulaic contributions and objectionable aspects of "social physics". @hansen1959 recast @eq-stewart-population-potential-sum to reflect accessibility, a model of human behaviour useful in capturing regularities in mobility patterns . In this equation, @hansen1959 replaced $M_j$ with _opportunities_ to derive _opportunity potential_, or more accurately, _potential of opportunities for interaction_ as follows:
$$
S_{i} = \sum_j O_j \cdot d_{ij}^\beta \\
$${#eq-accessibility}

@@

$S_{i}$ in @eq-accessibility is is a relative measure of the accessibility at $i$, $O_j$ is the size of the activity at $j$ and $d_{ij}$ is the distance or travel time between $i$ and $j$ with $\beta$ describing the effect size. Today, Hansen is is frequently cited as the father of modern accessibility analysis due to his work in @hansen1959 [e.g., @reggianiGuestEditorialNew2011], and the Hansen-type accessibility is commonly referred to today as the gravity-based accessibility measure.

Of note, however is that between @stewartDemographicGravitationEvidence1948 and @hansen1959 the proportionality constant $G$ in @eq-grav vanished.  There is some evidence that @hansen1959 was aware of the importance of this constant as he wrote about _direct_ and _inversely proportional_ relationships between opportunities in one area, opportunities at another, and their separation distance. Furthermore, those reading @hansen1959 could consult @stewartDemographicGravitationEvidence1948 in which the proportionality constant $G$  was set to 1 and that "$G$ [was] left for future determination: a suitable choice of other units can reduce it to unity".

@@

Since @hansen1959, accessibility analysis has been widely employed across planning disciplines, but to our knowledge, developments have not been concerned with explicitly determining a proportionality constant $G$. $G$ continues to be implicitly set to $1$ even when the fundamental relationship in accessibility is proportionality (e.g., $ S_{i}  \propto G \sum_j O_j \cdot f(d_{ij})$ and not equality [for instance, see the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020]. Without a proportionality constant, however, the units of $S_i$ remain unclear: the unit of 'potential of opportunity for interaction' is left free to change as $\beta$ is calibrated for the population center with no discussion of a constant to balance the units.

@@

However, it has been raised that there is a difficulty in interpreting gravity-based accessibility's meaning as anything other than a relative index of potential interaction (higher accessibility/lower accessibility) []. This interpretation plagues this type of analysis, as implications beyond high-or-low access hotspots are often desired. What does the accessibility value mean? Difficultly in interpreting the output's tangible meaning contributes to resistance to its application in practice [@ferreiraReenactingMobilityAccessibility2020]. As reviewed, quantity and variations on @eq-accessibility-general also compounds the issue of interpretability across studies.

@@

# Gravity-based accessibility's sibling: the gravity model

Today's gravity-based accessibility measure (@eq-accessibility-general) share the same early 20th century Newtonian-based  origins as the gravity model, the most widely used model in the study of spatial interaction [@ortuzar_2011_modelling;@wilson1971]. However, in some ways the gravity model and spatial interaction continued to evolve independently from @eq-accessibility-general, the former being developed into formalized transportation demand modelling practice to 'predict-and-provide', and the later taken up to describe land-use and transportation interactions by geography and social science- adjacent researchers and planners. <!-- Antonio, this is my sense at this point..?-->

There are plenty of insights to be shared across these sibling branches; they are both spatial interaction models From the perspective of gravity-based accessibility measure, this paper's focus is on the proportionality constant of the classic gravity model, which takes the following form:

$$
T_{ij} = k O_i D_j f(c_{ij})
$${#eq-gravity-model}

<!--Explain the features of the gravity model of today (Wilson) -->
Where $T_{ij}$ represents a $n$ x $m$ matrix of flows between an $n$ number of origins $i$ and an $m$ number of destinations $j$, $O_i$ and $D_j$ is a vector of mass attributes representing origin $i$ and destination $j$ respectively, and $f(c_{ij})$ is a $n$ x $m$ matrix representing some function of travel cost $c_{ij}$ which reflects travel impedance. In other words, $T_{ij}$ explicitly measures _interaction_ as a unit of trips and the single proportionality constant $k$ ensures the total sum of $T_{ij}$ represents the total flows in the data. It should be noted that the early analogy with the gravitational law proved too simplistic, and improvements to the gravity model (@eq-gravity-model) are the use of total trip ends (number of originating trips from zone $i$ are $O_i$ and ending trips at zone $j$ are $D_j$) instead of total populations and separation distance calibrated by a power $n$ (such as $P_i$, $P_j$, and $n$ in @eq-zipf, @eq-reilly and @eq-stewart-force). Finally, and of most important note in this paper, $k$ is a scale parameter that makes the overall equation proportional to the rate characteristic of the modeled phenomenon.

Traditionally, the development of the gravity model has put an emphasis on the interpretability of the resulting units [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967;@wilson1971]. As such, @eq-gravity-model can take account of any additional information that can be built into the model to constrain the _interaction_ variable $T_{ij}$. As mentioned, $T_ij$ is in the unit of trips, so additional knowledge can be the total trip flows $O_i$ originating at each $i$ and the total trip flows $D_j$ terminating at $j$. These pieces of information can constrain $T_{ij}$ as follows, ensuring it reflects this knowledge:

@@

Using these constraints, the gravity model @eq-gravity-model is typically distinguished in four cases, cases that are mathematically and conceptually consistent, so they have continued to be useful in classifying the gravity model in trip distribution modelling to this day [@ortuzar_2011_modelling]:

- neither @eq-constraint1-gravitymodel or @eq-constraint2-gravitymodel hold -- the unconstrained case;
- @eq-constraint1-gravitymodel holds -- the production constrained case (origin-constrained, a singly-constrained case); 
- @eq-constraint2-gravitymodel holds -- the attraction constrained case (destination-constrained, a singly-constrained case); 
- @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel both hold simultaneously -- the production-attraction constrained case (doubly-constrained).

@@

Referring to the preceding section, the entropy maximizing method can be used to derive an equivalent family of measures, but for the case of accessibility where the _interaction_ variable is $V_{ij}$, represent _potential interaction_, i.e., the number of potential reachable opportunities. $T_{ij}$ in @eq-gravity-model can be replaced with $V_{ij}$ and presented as follows:

@@

Like with @eq-gravity-model, we distinguish @eq-access into FIVE cases based on versions of the gravity-model building constraints (@eq-constraint1-gravitymodel,  @eq-constraint2-gravitymodel, and @eq-constraint3-gravitymodel). To inform each case of @eq-access with constraints that reflect knowledge of the macro-state, the total trip flow $O_i$ originating at each $i$ and the total trip flows $D_j$ terminating at $j$ are used. These pieces of information can constrain $V_{ij}$ as follows, ensuring it reflects this macro-level knowledge, i.e., the sum of the number of potential reachable opportunities.

>>>>

---
title: "A family of accessibility measures"
format: pdf
bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
---

@@

<!-- SUMMARY: Accessibility is related to mobility, however, institutionalized mobility-based transportation planning practice has treated accessibility as it's proxy. This has had problematic consequences. Conceptually, accessibility is a more complete and comprehensive evolution of mobility. As such, there has been recognition that the adoption of accessibility planning approaches are needed; however there are 1) too many definitions and 2) uncertainty in meaning of the outputs--> 
Accessibility, a central concept in transport geography, planning, and engineering, is conceptually related to human mobility and the opportunity landscape. In simple terms, it is defined as the "potential of opportunities for interaction" [@hansen1959]. The range of opportunities considered in accessibility analysis has grown to include <!-- --> employment [REFS], health care [REFS], green spaces [REFS], schools [REFS], social contacts [REFS], emergency services [e.g., shelters, REFS], and many more. In other words, accessibility analysis is today used to understand the potential to reach opportunities that are important to people [@ferreiraReenactingMobilityAccessibility2020]. Compared to earlier measures of performance used in transportation (e.g., VKT, PKT, etc.) that benchmark the amount of movement, accessibility brings a more holistic understanding of transportation and land use systems combined [@handyMeasuringAccessibilityExploration1997]. However, the focus of transportation planning has historically prioritized mobility and treated accessibility as a by-product of movement, leading to problematic consequences [@handy2020]. With the rise of an automobility mono-culture [@miller_collaborative_2011; @lavery_driving_2013], recent decades of mobility-based practice (e.g., road and highway expansion) have been marked by increased travel cost (time, effort, energy, financial cost), often with limited impact on accessibility levels [e.g., @handyACCESSIBILITYVSMOBILITYENHANCING2002; @paez_healthcare_2010]. In response, transportation planning practice has begin to adopt accessibility planning approaches, via accessibility planning tools [@silvaAccessibilityInstrumentsPlanning2017; @paez_developing_2013] or through the traditional four-step transportation demand model [@ortuzar_2011_modelling]. However, institutional and conceptual barriers remain, many of which relate to the diversity of accessibility definitions [@handy2020] and interpretability and communicability of the outputs [@geurs2004; @ferreiraReenactingMobilityAccessibility2020].

@@

<!-- PAPER'S AIM: improve interpretation of accessibility measure outputs. We do this by, 1) assuming accessibility is more complete understanding of mobiltiy, and connected them both mobility and accessibility are both based in the Netownian gravitational anology, lets introduce balancing factors like those  Gravity Model.   -->
In adopting the perspective that accessibility is an evolution of mobility conceptually, this paper aims to re-unify accessibility measures with its spatial interaction-based roots. It proposes to do so by mapping accessibility onto Wilson's family of spatial interactions models [@wilson1971]. In this, we aim not only to clarify accessibility's interpretation as a measure of performance, but also to encourage future developments that draw from its gravity modelling siblings, and spatial interaction history. To this end, we begin by tracing the development of spatial interaction from @ravensteinLawsMigration1889 to its split into accessibility research from @hansen1959 to today. After this, we derive different types of accessibility measures, following the family of gravity models namely: unconstrained, total constrained, origin- and destination- constrained, composite destination-constrained, and doubly constrained. Then, we will provide empirical examples of the types of measures. Finally, we will discuss the use of each measure and broadly the implications of ignoring the role of constrains in accessibility analysis.

@@

## The inquiry into spatial interaction traces to Newton's Universal Gravitation

The patterns of people's movement in space have been a subject of scientific inquiry for centuries. From as far back as Henry C. Carey's _Principles of Social Science_ [@careyPrinciplesSocialScience1858] there was already a concern with the scientific study of human interaction over space. It is in this work where Carey states that "man [is] the molecule of society [and their interaction is subject to] the direct ratio of the mass and the inverse one of distance" [@mckeanManual1883, pp. 37-38], which shows how investigations into human spatial interaction have often been explicitly coloured by the features of Newton's Law of Universal Gravitation [posited in _Principia Mathematica_ of 1687 and expressed in @eq-phys-grav-prop].

$$
F_{ij} \propto \frac{M_i M_j} {D_{ij}^{2}}
$${#eq-phys-grav-prop}

To be certain, the expression above, a proportionality, is one of the most famous in all of science. In brief, it states that the force of attraction $F$ between a pair of bodies $i$ and $j$ is directly _proportional_ to the product of their masses $M_i$ and $M_j$, and inversely _proportional_ to the square of the distance between them $D_{ij}$. Direct proportionality only means that as the product of the masses increases, so does the force. Likewise, inverse proportionality only means that as the distance increases, the force decreases. To quantify the magnitude of the force, an empirical factor, say $G$, is required to convert the proportionality into an equality, which ensures that the force $F$ matches empirical observations of the force of attraction between masses. Ultimately, the equation for the force is as seen in @eq-phys-grav, where $G$ is a proportionality constant obtained empirically:

$$
F_{ij} = G \frac{M_i M_j} {D_{ij}^{2}}
$${#eq-phys-grav}

Newton's initial estimate of $G$ was based on a speculation that the mean density of earth was between five or six times that of water, an assumption that Hutton's experiments of 1778 validated [@hutton_xxxiii_1778, p. 783]. Still, it took over a century from the publication of _Principia_ to refine the estimate of this constant to within 1\% accuracy, with Cavendish's 1798 experiment [@cavendish_xxi_1798].

## Early research on human spatial interaction: from Ravenstein (1889) to Stewart (1948)

<!-- SUMMARY: from the 1880s to the 1940s, researchers have theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance. This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy. -->
Following Carey's _Principles_ of 1858, research into human spatial interaction continued in different contexts. In the late 1880s Ravenstein proposed his "Laws of Migration", based on his empirical analysis of migration flows in various countries [@ravensteinLawsMigration1885; @ravensteinLawsMigration1889]. In these works, Ravenstein posited a directly proportional relationship of migration flows with the size of the destination (i.e., centres of commerce and industry), and an inversely proportional relationship between the size of flows and separation between origin and destination. These propositions are similar to Newton's gravitational laws. Over time, other researchers discovered similar relationships. <!--Rilley does not cite Newton, nor Carey, nor Ravenstein, but I removed this curiosity as it's not central to our point here--> For example, @reilly1929methods formulated a law of retail gravitation, expressed in terms of equal attraction to competing retail destinations. Later, Zipf proposed a $\frac{P_1P_2}{D}$ hypothesis for the case of goods movement by railways @zipfHypothesisCaseRailway1946, intercity personal movement @zipfHypothesisIntercityMovement1946,and information @zipfDeterminantsCirculationInformation1946. The $\frac{P_1P_2}{D}$ hypothesis stated that the magnitude of flows was proportional to the product of the populations of settlements, and inversely proportional to the distance between them.

A common theme among these early researchers of human spatial interaction is that a proportionality constant similar to $G$ in @eq-phys-grav is missing. Of the researchers cited, only Reilly and Zipf expressed their hypotheses in mathematical terms. Reilly's hypothesis was expressed as follows:
$$
B_a = \frac{(P_a\cdot P_T)^N}{D_{aT}^n}
$${#eq-reilly}

@@

After Carey, it is in Stewart's work on the principles of demographic gravitation that we find the strongest connection yet to Newton's law [@stewartDemographicGravitationEvidence1948]. We suspect that this may relate to academic backgrounds; Stewart was a physicist, whereas Ravenstein, Reilly, and Zipf were social scientists. Besides awareness of preceding research, as he cites Reilly and Zipf as predecessors in the analysis of human spatial interaction, Stewart appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, as follows:
$$
F = G\frac{(N_1\mu_1)(N_2\mu_2)}{d_{12}^2}
$${#eq-stewart-force}

@@

Stewart's formulation of demographic force, developed in the context of what he called "social physics" [@stewartPrinciples1947], was problematic. It had issues with inconsistent mathematical notation. For example, e.g., in @stewartDemographicGravitationEvidence1948, $G$ is used as a proportionality constant for both _demographic force_ and  _demographic energy_ on p. 34 <!--Antonio, is this what you meant?-->. Perhaps more seriously, Stewart's work was permeated by a view of humans as particles following physical laws, but with ideas that were afflicted by unscientific racism. For instance, the molecular weight $\mu$ was presumed to be one for the average American, but "presumably...much less than one....for an Australian aborigine" [p. 35]. Stewart's ideas about "social physics" fell out of favour among social scientists, but not before influencing the emerging field of accessibility research.

In summary: from the 1850s to the 1940s, researchers attempted to theoretically and empirically characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $M$ of a pair of locations ($i$ and $j$) and inversely proportional by some sort of separation distance $D_{ij}$. Those that captured the relationship empirically, approached the concept with different expressions, some including a proportionality constant and some not, but all tying back to the Newtonian gravity analogy.

@@

From @stewartDemographicGravitationEvidence1948, we arrive to 1959 and Walter G. Hansen, whose work proved to be exceptionally influential in the accessibility literature [@hansen1959]. In his seminal paper, Hansen defined accessibility, i.e., "the potential of opportunities for interaction" [p.73], as "a generalization of the population-over-distance relationship or 'population potential' concept developed by Stewart [@stewartDemographicGravitationEvidence1948]" [p.73]. As well as being a student of city and regional planning at Massachusetts Institute of Technology, Hansen was also an engineer with the the Bureau of Roads, and likely preoccupied with applied research into the power of transportation to shape land uses. @hansen1959 focused on @stewartDemographicGravitationEvidence1948's _population potential_ (expressed in @eq-stewart-population-potential-sum), leaving Stewart's other formulaic contributions and objectionable aspects of "social physics" behind. @hansen1959 recast @eq-stewart-population-potential-sum to reflect accessibility, a model of human behaviour useful to capture regularities in mobility patterns . In this equation, @hansen1959 replaced $M_j$ with _opportunities_ to derive an _opportunity potential_, or more accurately, a _potential of opportunities for interaction_ as follows:
$$
S_{i} = \sum_j O_j \cdot d_{ij}^\beta \\
$${#eq-accessibility}

@@

$S_{i}$ in @eq-accessibility is a measure of the accessibility at $i$, whereas $O_j$ is the size of the activity at $j$ and $d_{ij}$ is the distance or travel time between $i$ and $j$ with $\beta$ describing the effect size. Today, Hansen is is frequently cited as the father of modern accessibility analysis [e.g., @reggianiGuestEditorialNew2011], and Hansen-type accessibility is commonly referred to today as the gravity-based accessibility measure.

Of note, however is that between @stewartDemographicGravitationEvidence1948 and @hansen1959 the proportionality constant $G$ in @eq-stewart-population-potential vanished. There is some evidence that @hansen1959 was aware of the importance of this constant as he wrote about _direct_ and _inversely proportional_ relationships between opportunities in one area, opportunities at another, and their separation distance. Furthermore, those reading @hansen1959 should remember that @stewartDemographicGravitationEvidence1948 set the proportionality constant $G$  to 1, with a note that"$G$ [was] left for future determination: a suitable choice of other units can reduce it to unity".

@@

Since @hansen1959, accessibility analysis has been widely employed across planning disciplines [CITE MAJOR REVIEWS OVER LAST 30 YEARS]. However, to our knowledge, the proportionality constant has remained forgotten, with no notable developments to explicitly determining it. In this way, $G$ continues to be implicitly set to $1$ even when the fundamental relationship in accessibility is proportionality (e.g., $ S_{i}  \propto G \sum_j O_j \cdot f(d_{ij})$ and not equality [for instance, see the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020]. Without a proportionality constant, however, the units of $S_i$ remain unclear: the unit of 'potential of opportunity for interaction' is left free to change as $\beta$ is calibrated for the population center with no discussion of a constant to balance the units.

@@

However, it has been raised that there is a difficulty in interpreting gravity-based accessibility's meaning as anything other than a relative index of potential interaction (higher accessibility/lower accessibility) []. Interpretability issues plague this type of analysis, as implications beyond high-or-low access hotspots are often desired. What does the accessibility value mean? Difficultly in interpreting the output's tangible meaning contributes to resistance to its application in practice [@ferreiraReenactingMobilityAccessibility2020]. As reviewed, quantity and variations on @eq-accessibility-general also compounds the issue of interpretability across studies.

@@

# Gravity-based accessibility's sibling: Wilson's spatial interaction model

Today's gravity-based accessibility measures (@eq-accessibility-general) share the same early 20th century Newtonian-based origins as the model's of Reilly, Zipf, and Stewart, versions of which continue to be extensively used [@ortuzar_2011_modelling;@wilson1971]. However, in some ways the gravity model and spatial interaction continued to evolve independently from @eq-accessibility-general, the former being developed into formalized transportation demand modelling practice to 'predict-and-provide', and the later taken up to describe land-use and transportation interactions by geography and social science-adjacent researchers and planners. <!-- Antonio, this is my sense at this point..?-->

There are plenty of insights to be shared across these sibling branches; they are both spatial interaction models From the perspective of gravity-based accessibility measure, this paper's focus is on the proportionality constant of the spatial interaction model, which takes the following form after Wilson [-@wilson1971]:

$$
T_{ij} = k O_i D_j f(c_{ij})
$${#eq-phys-gravity-model}

<!--Explain the features of the gravity model of today (Wilson) -->
Where $T_{ij}$ represents a $n$ x $m$ matrix of flows between an $n$ number of origins $i$ and an $m$ number of destinations $j$, $O_i$ and $D_j$ is a vector of mass attributes representing origin $i$ and destination $j$ respectively, and $f(c_{ij})$ is a $n$ x $m$ matrix representing some function of travel cost $c_{ij}$ which reflects travel impedance. In other words, $T_{ij}$ explicitly measures _interaction_ as a unit of trips and the single proportionality constant $k$ ensures the total sum of $T_{ij}$ represents the total flows in the data. It should be noted that the early analogy with the gravitational law proved too simplistic, and improvements to the gravity model (@eq-phys-gravity-model) are the use of total trip ends (number of originating trips from zone $i$ are $O_i$ and ending trips at zone $j$ are $D_j$) instead of total populations and separation distance calibrated by a power $n$ (such as $P_i$, $P_j$, and $n$ in @eq-zipf, @eq-reilly and @eq-stewart-force). Finally, and of most important note in this paper, $k$ is a scale parameter that makes the overall equation proportional to the rate characteristic of the modeled phenomenon.

Traditionally, the development of the gravity model has put an emphasis on the interpretability of the resulting units [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967;@wilson1971]. As such, @eq-phys-gravity-model can take account of any additional information that can be built into the model to constrain the _interaction_ variable $T_{ij}$. As mentioned, $T_ij$ is in the unit of trips, so additional knowledge can be the total trip flows $O_i$ originating at each $i$ and the total trip flows $D_j$ terminating at $j$. These pieces of information can constrain $T_{ij}$ as follows, ensuring it reflects this knowledge:

Total constrained:
$$
\sum_i\sum_j T_{ij} = T
$${#eq-constraint0-gravitymodel}

@@

```{r}
T_1 <- matrix(c(95, 23, 42, 27, 378, 45, 38, 36, 106), ncol = 3, byrow = TRUE)

T_1
```

```{r}
C <- matrix(c(2, 15, 5, 15, 2, 10, 5, 10, 2), ncol = 3, byrow = TRUE)

C
```

```{r}
O_i <- rowSums(T_1)
D_j <- colSums(T_1)

O_i
D_j
```

```{r}
T <- sum(sum(T_1))

T
```

Unconstrained (note that there is no proportionality constant!):
$$
\hat{T_{ij}} = \frac{O_i\cdot D_j}{C_ij}
$$

```{r}
T_unc <- O_i %*% t(D_j) / C

T_unc
```

If we do the sums over rows, columns and all:
```{r}
rowSums(T_unc)
colSums(T_unc)
sum(sum(T_unc))
```

Total flow constrained (needs a single/global proportionality constant):
$$
\hat{T_{ij}} = k\frac{O_i\cdot D_j}{C_{ij}}
$$

```{r}
k <- T/sum(sum(T_unc))
k
```

```{r}
T_tot <- k * T_unc

T_tot
```

Let's check those marginals and total:
```{r}
rowSums(T_tot)
colSums(T_tot)
sum(sum(T_tot))
```

Destination constrained:
$$
\hat{T_{ij}} = \frac{B_j\cdot O_i\cdot D_j}{C_{ij}}
$$
```{r}
B_j <- colSums(T_1)/colSums(T_unc)

B_j <- matrix(rep(B_j, 3), ncol = 3, byrow = TRUE)

B_j
```

```{r}
T_dest <- B_j * T_unc

T_dest
```

Let's check those marginals and total:
```{r}
rowSums(T_dest)
colSums(T_dest)
sum(sum(T_dest))
```

Origin constrained:
$$
\hat{T_{ij}} = \frac{A_i\cdot O_i\cdot D_j}{C_{ij}}
$$

```{r}
A_i <- rowSums(T_1)/rowSums(T_unc)

A_i <- matrix(rep(A_i, 3), ncol = 3, byrow = FALSE)

A_i
```

```{r}
T_orig <- A_i * T_unc

T_orig
```

Let's check those marginals and total:
```{r}
rowSums(T_orig)
colSums(T_orig)
sum(sum(T_orig))
```

Now how does this translate to accessibility?

Opportunities (say, grocery stores):
```{r}
D_j <- c(7, 3, 2)
```

Unconstrained accessibility (no proportionality constant):
$$
S_i = \sum_j\frac{D_j}{C_{ij}}
$$

```{r}
S_i <- (1/C) %*% D_j
```

These numbers are proportional to the accessibility (i.e., higher or lower), but are not scaled to number of stores. Check the sum:
```{r}
sum(S_i)
```

Total flow constrained accessibility, needs a balancing factor:
```{r}
k <- sum(D_j)/sum(S_i)

k
```

Total flow constrained accessibility:
Unconstrained accessibility (no proportionality constant):
$$
S_i = k\sum_j\frac{D_j}{C_{ij}}
$$

```{r}
S_tot <- k * S_i

S_tot
```

Now check the sum:
```{r}
sum(S_tot)
```

Using these constraints, the gravity model @eq-phys-gravity-model is typically distinguished in four cases, cases that are mathematically and conceptually consistent, so they have continued to be useful in classifying the gravity model in trip distribution modelling to this day [@ortuzar_2011_modelling]:

- None of @eq-constraint0-gravitymodel, @eq-constraint1-gravitymodel, and @eq-constraint2-gravitymodel hold -- the unconstrained case;
- Only @eq-constraint0-gravitymodel holds -- total flow constrained
- @eq-constraint1-gravitymodel holds -- the production constrained case (origin-constrained, a singly-constrained case); 
- @eq-constraint2-gravitymodel holds -- the attraction constrained case (destination-constrained, a singly-constrained case); 
- @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel both hold simultaneously -- the production-attraction constrained case (doubly-constrained).

@@

Referring to the preceding section, the entropy maximizing method can be used to derive an equivalent family of measures, but for the case of accessibility where the _interaction_ variable is $V_{ij}$, represent _potential interaction_, i.e., the number of potential reachable opportunities. $T_{ij}$ in @eq-phys-gravity-model can be replaced with $V_{ij}$ and presented as follows:

@@

Like with @eq-phys-gravity-model, we distinguish @eq-access into FIVE cases based on versions of the gravity-model building constraints (@eq-constraint1-gravitymodel,  @eq-constraint2-gravitymodel, and @eq-constraint3-gravitymodel). To inform each case of @eq-access with constraints that reflect knowledge of the macro-state, the total trip flow $O_i$ originating at each $i$ and the total trip flows $D_j$ terminating at $j$ are used. These pieces of information can constrain $V_{ij}$ as follows, ensuring it reflects this macro-level knowledge, i.e., the sum of the number of potential reachable opportunities.
